Finally Tagless Staged Interpreters for Simpler Typed Languages

### 2007-04-06 tagless/tagless-final.tex
Done with Section 5

The module returned by |CPST| does not define |repr|
and thus does not have signature |Symantics|.
The reason is again the result type of |lam f|. Whereas
|int i| or |add e1 e2| have the (abbreviated) 
type |('c,..., (int -> 'w) -> 'w) S.repr|,
the result of |lam (fun x -> add x (int 1))| has type\\
|('c,...((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr|. 
Hence, to write the type equation for |CPST.repr| we again need a 
type-discriminating
function similar to the type function |repr_pe| of
\S\ref{S:PE-solution}. Alas, the function we need here is
similar but not identical to |repr_pe|. Thus we need to add 
another type variable to the |repr| type and modify the Symantics
signature accodringly. Again, the terms in previous Symantics module
stay unchanged, but the |repr| type equations in those structures have to
account for a new (phantom) type variable.

To save space, we show a simplified and less principled approach that
lets us use the module |CPST| as is. Because the module does not
have the signature |Symantics|, we cannot apply the |EX| functor to it.
But we can nevertheless write the tests:
\begin{code}
module T = struct
 module M = CPST(C)
 open M
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = ... (* the same as before *)
 let testpowfix7 = (* the same as before *)
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
That is, we instantiate |CPST| with the desired base intrepreter (the
compiler |C| in the case above) so that we can use the result |M| to
interpret object level terms. Those terms are \emph{exactly} as before.
Having to textually copy the terms is the
price we pay for this simplified treatment. We shall also encounter this
copying when discussing self-interpretation.  We
will see then that the copying step is not frivolous, but represents
a profound instance of 
``plugging the hole'', which is one of the many faces of polymorphism.

It is instructive to evaluate a few examples. With 
|CPST| instantiated with the compiler~|C| above,
|T.test1| gives
\begin{code}
.<fun x_5 -> ((fun x_2 -> 
        (x_2 (fun x_3 -> fun x_4 -> (x_4 x_3))))
        (fun x_6 -> ((fun x_1 -> (x_1 (true))) 
        (fun x_7 -> ((x_6 x_7) x_5)))))>.
\end{code}
which shows the na\"{\i}ve CPS transformation of the original code.
The result is sub-optimal with several apparent $\beta$-redexes. 
By instantiating |CPST| with |P| instead, we get the expected
\begin{code}
  P.repr = {P.st = Some <fun>;
            P.dy = .<fun x_5 -> (x_5 (true))>.}
\end{code}

>>>>

The module returned by |CPST| does not define |repr|
and thus does not have signature |Symantics|.
The reason is again the type of |lam f|. Whereas
|int| and |add| return the (abbreviated) type
|('c, ..., (int -> 'w) -> 'w) S.repr|,
the type of |lam (add (int 1))| is
\begin{code}
('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2)
          -> 'w2) S.repr
\end{code}
Hence, to write the type equation defining |CPST.repr| we again need
a type function with a typecase distinction, similar to |repr_pe|
in~\S\ref{S:PE-solution}. Alas, the type function we need is not
identical to |repr_pe|, so we need to add another type argument to
|repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
terms in previous implementations of |Symantics| stay unchanged, but the
|repr| type equations in those implementations have to take a new
(phantom) type argument.

To save space, we just
use the module returned by |CPST| as is. Because it does not
match the signature |Symantics|, we cannot apply the |EX| functor to it.
Nevertheless, we can write the tests.
\begin{code}
module T = struct
 module M = CPST(C)
 open M
 let test1 () = (* same as before *)
   app (lam (fun x -> x)) (bool true)
 let testpowfix () = ... (* same as before *)
 let testpowfix7 = (* same as before *)
    lam (fun x -> app (app (testpowfix ()) x)
                      (int 7))
end
\end{code}
We instantiate |CPST| with the desired base interpreter~|C|,
then use the result |M| to
interpret object terms. Those terms are \emph{exactly} as before.
Having to textually copy the terms is the
price we pay for this simplified treatment.
Our discussion of self\hyp interpretation in~\S\ref{selfinterp} shows
that this copying is not frivolous but represents plugging a term into
a context, which is one of the many faces of polymorphism.

With 
|CPST| instantiated by the compiler~|C| above,
|T.test1| gives
\begin{code}
.<fun x_5 ->
  (fun x_2 -> x_2 (fun x_3 x_4 -> x_4 x_3))
  (fun x_6 -> (fun x_1 -> x_1 true)
              (fun x_7 -> x_6 x_7 x_5))>.
\end{code}
This output is the na\"{\i}ve CPS transformation of $(\fun{x}x)\True$,
containing several apparent $\beta$-redexes.  To eliminate these
redexes, we simply change~|T| to instantiate |CPST| with |P| instead.
\begin{code}
{P.st = Some <fun>;
 P.dy = .<fun x_5 -> x_5 true>.}
\end{code}
